---
title: "Storm Data Analysis"
author: "Chaitanya R"
date: "December 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```

## Synopsis

U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database is available for public use. This database has information on major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Objective of the document is to explore which types of weather events cause the most amount of damage to public health and/or economy. Secondary objective is to practice data exploration using R. 

Since historical data is sparsely polulated, records before 1996 were ignored. One of the key variables, EVTYPE, has been inconsistently coded across years. Effort has been spent to clean this variable up. Simply summarising the FATALITIES and INJURIES variables gave us answers to the first part of our primary objective. For estimating the economic damage, both property and crop damage were considered and adjusted based on the units provided in another variable. Finally total damage is computed, which was summarised by the weather events to answer the second question of interest.

## Data Download

Data was downloaded from internet(a compressed csv file), and loaded into a data frame.

```{r cache=TRUE}
# download the data file
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","StormData.csv.bz2")

# Read it into data frame 
StormData <- read.csv("StormData.csv.bz2")
```

## Data Processing

First, some basic checks - What are the columns in the dataset?

```{r}
names(rawdata)
```

We might not need all 37 columns for our analysis. Hence, shortlist only a few required variables: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. Also, since historical data is sparsely polulated, records before 1996 were ignored. These two will help in speeding up the further steps.

```{r}
library(dplyr)

StormData2 <- select(StormData, BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
# Format the date column as date and filter data from 1996 and later
StormData2$BGN_DATE <- as.Date(StormData2$BGN_DATE, format = "%m/%d/%Y")
StormData2 <- filter(StormData2, format(BGN_DATE, "%Y") >= 1996)

```

Now, EVTYPE is the key variable of interest. This has been inconsistently coded across years. The following effort was spent to clean this variable up.


```{r}

# Clean the variable for typos
# remove leading and traling spaces and Convert all to Upper case
StormData2$EVTYPE <- toupper(trimws(StormData2$EVTYPE))
# Remove multiple spaces
StormData2$EVTYPE <- gsub("[[:space:]]+", " ", StormData2$EVTYPE)
# Other typo substitutions
StormData2$EVTYPE <- gsub("EXTREMELY|EXCESSIVE|EXCESSIVELY", "EXTREME", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("FLOODINGINGS*|FLOOODING", "FLOOD", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("LIGHTNING\\.|LIGHTING|LIGNTNING", "LIGHTNING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("NON-", "NON ", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("PROLONG", "PROLONGED", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RAINS", "RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RIP CURRENTS", "RIP CURRENT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("STORMS", "STORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TORNDAO|TORNADOES", "TORNADO", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TSTM|TH*UND*ER*[A-Z]*RMW*|THUNDERSTROM|THUDERSTORM",
                    "THUNDERSTORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("UNUSUALLY", "UNUSUAL", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WILD.*FIRE.*|WILD/FOREST.*", "WILD/FOREST FIRES", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WINDS|WND", "WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WINTERY", "WINTER", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WARMTH", "WARM", StormData2$EVTYPE)
# grouping some events
StormData2$EVTYPE <- gsub("^BLIZZARD.*|ICE STORM", "BLIZZARD", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^COASTAL.*|.*/CSTL .*", "COASTAL EROSION/FLOODING/STORM ",
                    StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("EXTREME COLD.*|EXTENDED COLD.*", "EXTREME COLD", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^DRY.*", "DRY CONDITIONS", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^FLOODING.*", "FLOODING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^FREEZE|^FREEZING.*|^FROST.*",
                    "FREEZING FOG/RAIN/SLEET/SNOW", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HAIL.*", "HAIL", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("DROUGHT|EXTREME HEAT.*|^HEAT.*", "EXTREME HEAT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HEAVY RAIN.*", "HEAVY RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HURRICANE.*", "HURRICANE", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HEAVY SNOW.*|^SNOW.*|EXCESSIVE SNOW", "HEAVY SNOW/ICE",
                    StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("LIGHTNING.*", "LIGHTNING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^MARINE.*", "MARINE THUNDERSTORM/ACCIDENT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RAIN.*|PROLONGEDED RAIN", "RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RIP CURRENT.*|HEAVY SURF.*|HIGH SURF.*", "HIGH SURF", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("SLEET.*", "SLEET", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("VOLCANIC.*", "VOLCANIC", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("THUNDERSTORM.*|SEVERE THUNDERSTORM", "THUNDERSTORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TORNADO.*", "TORNADO", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TROPICAL STORM.*", "TROPICAL STORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("UNSEASONAL.*|^UNSEASONABL[EY].*|^^UNUSUAL.*",
                    "UNUSUAL WEATHER CONDITION", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HIGH WIND.*|STRONG WIND.*|^WIND.*", "HIGH WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^WATERSPOUT.*|WATER SPOUT", "WATERSPOUT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^WINTER.*", "WINTER STORM/WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^NONE|^SUMMARY.*", "?", StormData2$EVTYPE)

```

# Summarising Fatalities and Injuries

Fatalities and Injuries are the two sets of information available to estimate the impact of the weather events on public health. To sum up the total fatalaties by event type, and taking the top 10 events, the following code was used. Injuries was also analysed similarly.

```{r}
Top10FatByEvnt <- select(StormData2, EVTYPE, FATALITIES) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(FATALITIES)) %>%
    slice(1:10)

Top10InjByEvnt <- select(StormData2, EVTYPE, INJURIES) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(INJURIES)) %>%
    slice(1:10)

```

These may be plotted for more clarity:

```{r}
# Plots for Human injuries
library(ggplot2)
ggplot(Top10FatByEvnt, aes(x=EVTYPE, y=FATALITIES)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("FATALITIES") + ggtitle("Top 10 Events by FATALITIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 
ggplot(Top10InjByEvnt, aes(x=EVTYPE, y=INJURIES)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("INJURIES") + ggtitle("Top 10 Events by INJURIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

```

# Economic damage

For estimating economic damage, property damage and crop damage information was provided. Each of this is coded in different units which is avaiable in EXP ending columns. Total damage was computed as shown in the code that follows. 

Summaring the damage and plotting it is done in a similar fashion as in the previous section.

```{r}
StormData2$PROPDMGEXP <- toupper(trimws(StormData2$PROPDMGEXP))
# Standardising all the data in '000s 
StormData2$PROP_DMG_K <- StormData2$PROPDMG*ifelse(StormData2$PROPDMGEXP =="K",1,ifelse(StormData2$PROPDMGEXP =="M",1000,ifelse(StormData2$PROPDMGEXP =="B",1e+06,ifelse(StormData2$PROPDMGEXP == "H",0.1,0.001))))
StormData2$CROP_DMG_K <- StormData2$CROPDMG*ifelse(StormData2$CROPDMGEXP =="K",1,ifelse(StormData2$CROPDMGEXP =="M",1000,ifelse(StormData2$CROPDMGEXP =="B",1e+06,ifelse(StormData2$CROPDMGEXP == "H",0.1,0.001))))
# Total damage is property damage + Crop damage
StormData2$TOT_DMG_K <- StormData2$PROP_DMG_K + StormData2$CROP_DMG_K

Top10DmgByEvnt <- select(StormData2, EVTYPE, TOT_DMG_K) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(TOT_DMG_K)) %>%
    slice(1:10)

ggplot(Top10DmgByEvnt, aes(x=EVTYPE, y=TOT_DMG_K)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("TOT_DMG_K") + ggtitle("Top 10 Events by Economic Damage") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

```

