---
title: "Storm Data Analysis"
author: "Chaitanya R"
date: "December 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
```

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This Coursera project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

The basic goal of this assignment is to explore the NOAA Storm Database and answer two basic questions about severe weather events: across the United States, which types of events (1) are most harmful with respect to population health and (2) have the greatest economic consequences.

## Data Processing

First we download from internet the data (a compressed csv file), and load the dataset into memory.



```{r cache=TRUE}
# download the data file
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","StormData.csv.bz2")
# Read it into data frame 
StormData <- read.csv("StormData.csv.bz2")
```

We can now have a look at the variable names of the NOAA Storm Database:


```{r echo=FALSE}
names(StormData)
```

For our analysis we only need a few variables, namely: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. We create a new smaller dataset containing this subset to speed up the further data manipulation process.

```{r}
library(dplyr)

StormData2 <- select(StormData, BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
# Format the date column as date and filter data from 1996 and later
StormData2$BGN_DATE <- as.Date(StormData2$BGN_DATE, format = "%m/%d/%Y")
StormData2 <- filter(StormData2, format(BGN_DATE, "%Y") >= 1996)

```

The variable EVTYPE contains all sort of types of weather events with case combinations of strings variable across the years. Make all the letters uppercase to fix this point.


```{r}

# Clean the variable for typos
# remove leading and traling spaces and Convert all to Upper case
StormData2$EVTYPE <- toupper(trimws(StormData2$EVTYPE))
# Remove multiple spaces
StormData2$EVTYPE <- gsub("[[:space:]]+", " ", StormData2$EVTYPE)
# Other typo substitution
StormData2$EVTYPE <- gsub("EXTREMELY|EXCESSIVE|EXCESSIVELY", "EXTREME", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("FLOODINGINGS*|FLOOODING", "FLOOD", StormData2$EVTYPE)
# StormData2$EVTYPE <- gsub("FLASH FLOODING.*", "FLOODING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("LIGHTNING\\.|LIGHTING|LIGNTNING", "LIGHTNING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("NON-", "NON ", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("PROLONG", "PROLONGED", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RAINS", "RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RIP CURRENTS", "RIP CURRENT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("STORMS", "STORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TORNDAO|TORNADOES", "TORNADO", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TSTM|TH*UND*ER*[A-Z]*RMW*|THUNDERSTROM|THUDERSTORM",
                    "THUNDERSTORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("UNUSUALLY", "UNUSUAL", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WILD.*FIRE.*|WILD/FOREST.*", "WILD/FOREST FIRES", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WINDS|WND", "WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WINTERY", "WINTER", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("WARMTH", "WARM", StormData2$EVTYPE)
# grouping some events
StormData2$EVTYPE <- gsub("^BLIZZARD.*|ICE STORM", "BLIZZARD", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^COASTAL.*|.*/CSTL .*", "COASTAL EROSION/FLOODING/STORM ",
                    StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("EXTREME COLD.*|EXTENDED COLD.*", "EXTREME COLD", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^DRY.*", "DRY CONDITIONS", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^FLOODING.*", "FLOODING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^FREEZE|^FREEZING.*|^FROST.*",
                    "FREEZING FOG/RAIN/SLEET/SNOW", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HAIL.*", "HAIL", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("DROUGHT|EXTREME HEAT.*|^HEAT.*", "EXTREME HEAT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HEAVY RAIN.*", "HEAVY RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HURRICANE.*", "HURRICANE", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HEAVY SNOW.*|^SNOW.*|EXCESSIVE SNOW", "HEAVY SNOW/ICE",
                    StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("LIGHTNING.*", "LIGHTNING", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^MARINE.*", "MARINE THUNDERSTORM/ACCIDENT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RAIN.*|PROLONGEDED RAIN", "RAIN", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("RIP CURRENT.*|HEAVY SURF.*|HIGH SURF.*", "HIGH SURF", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("SLEET.*", "SLEET", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("VOLCANIC.*", "VOLCANIC", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("THUNDERSTORM.*|SEVERE THUNDERSTORM", "THUNDERSTORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TORNADO.*", "TORNADO", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("TROPICAL STORM.*", "TROPICAL STORM", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("UNSEASONAL.*|^UNSEASONABL[EY].*|^^UNUSUAL.*",
                    "UNUSUAL WEATHER CONDITION", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("HIGH WIND.*|STRONG WIND.*|^WIND.*", "HIGH WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^WATERSPOUT.*|WATER SPOUT", "WATERSPOUT", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^WINTER.*", "WINTER STORM/WIND", StormData2$EVTYPE)
StormData2$EVTYPE <- gsub("^NONE|^SUMMARY.*", "?", StormData2$EVTYPE)

```


```{r}
Top10FatByEvnt <- select(StormData2, EVTYPE, FATALITIES) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(FATALITIES)) %>%
    slice(1:10)

Top10InjByEvnt <- select(StormData2, EVTYPE, INJURIES) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(INJURIES)) %>%
    slice(1:10)

```



```{r}
# Plots for Human injuries
library(ggplot2)
ggplot(Top10FatByEvnt, aes(x=EVTYPE, y=FATALITIES)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("FATALITIES") + ggtitle("Top 10 Events by FATALITIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 
ggplot(Top10InjByEvnt, aes(x=EVTYPE, y=INJURIES)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("INJURIES") + ggtitle("Top 10 Events by INJURIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

```



```{r}
StormData2$PROPDMGEXP <- toupper(trimws(StormData2$PROPDMGEXP))
StormData2$PROP_DMG_K <- StormData2$PROPDMG*ifelse(StormData2$PROPDMGEXP =="K",1,ifelse(StormData2$PROPDMGEXP =="M",1000,ifelse(StormData2$PROPDMGEXP =="B",1e+06,ifelse(StormData2$PROPDMGEXP == "H",0.1,0.001))))
StormData2$CROP_DMG_K <- StormData2$CROPDMG*ifelse(StormData2$CROPDMGEXP =="K",1,ifelse(StormData2$CROPDMGEXP =="M",1000,ifelse(StormData2$CROPDMGEXP =="B",1e+06,ifelse(StormData2$CROPDMGEXP == "H",0.1,0.001))))
StormData2$TOT_DMG_K <- StormData2$PROP_DMG_K + StormData2$CROP_DMG_K

Top10DmgByEvnt <- select(StormData2, EVTYPE, TOT_DMG_K) %>%
    group_by(EVTYPE) %>%
    summarise_all(funs(sum)) %>%
    arrange(desc(TOT_DMG_K)) %>%
    slice(1:10)

ggplot(Top10DmgByEvnt, aes(x=EVTYPE, y=TOT_DMG_K)) + 
    geom_bar(aes(x=EVTYPE), stat="identity") + 
    xlab("EVTYPE") + ylab("TOT_DMG_K") + ggtitle("Top 10 Events by Economic Damage") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

```

